---
title: "English Language Dictionary"
output: html_notebook
---

```{r}
library(dplyr)
```

This notebook contains the process to build an english language dictionary for comparison of words and construction of ngram models in this package. The source data used by this project is the [english-words](https://github.com/dwyl/english-words) project, which reportedly obtained its data from InfoChimps, although the link is broken as of this writing.

We downloaded the most recent word file and saved it into our raw data as `en_words.txt`. To be more specific, we downloaded [this](https://github.com/dwyl/english-words/blob/a74d7b5bc6d7f9527f079a8f0fbf38695613745c/words3.txt) version of file.

```{r raw-data}
en_words = readLines(file.path('..', 'data-raw', 'en_words.txt'))
```

With the raw data in hand, we now need to make our dictionary match our ngram table. That means performing the same preprocessing steps. However, we can't do this without some consideration to what our processing will do to the dictionary. If certain words 

```{r corpus}
library(quanteda)
en_corp = corpus(en_words)
en_mat = dfm(
  corp,
  ngrams=1, tolower=TRUE,
  removeNumbers = TRUE, removePunct = TRUE,
  removeSymbols = TRUE
)
```

```{r}
"Th5 7uIc# *ro@n 9(x" %>% 
  quanteda::tokens(
    what="fastestword",
    removeNumbers=TRUE, removePunct=TRUE, removeSymbols=TRUE
  ) %>% .[[1]]
```

